Nucleotide Diversity Loss on a Plant Y Chromosome Following Recent Recombination Suppression

### 2016-11-09 manuscript/manuscript.tex
more intro edits

One fundamental difference between the X and Y chromosomes is that there are 1/3 as many Y-linked gene copies as X-linked ones in a diploid population. Because of this, the effective population size ($N_{e}$) of the Y chromosome is expected to be 1/4 that of the autosomes, whereas the $N_{e}$ for the X chromosome should be 3/4 that of autosomes (assuming an equal number of reproducing females and males). The lowered $N_{e}$ of the Y chromosome implies that deleterious mutations are more likely to rise to high frequency because of genetic drift \citep{nei1981genetic}. Moreover, this loss of recombination on the Y implies that genes on this chromosome will experience strong linkage disequilibrium, making them vulnerable to "Hill-Robertson Interference", in which the genetic associations between selected alleles and their genetic backgrounds cause a reduction in the efficacy of selection \citep{fisher1930genetical, muller1964relation, hill1966HReffect, mcvean2000}. This suggests that Y chromosomes should harbor a lower amount of neutral genetic variability than predicted based on the number of Y chromosomes in a population, though the extent to which selective interference should affect chromosome-wide levels of nucleotide diversity is not known. However, given that a large number of sites are probably under weak selection (refs), the "interference" effect might alone cause a substantial diversity loss  important feature of the evolution of nonrecombining genomes.

(though simulations suggest a reduction of 30 \% relative to neutrality when there are 100,000 completely linked sites.

which at equilibrium is proportional to the product of $N_{e}$ and the mutation rate, $\mu$ \citep{Kimura1984,charlesworth1987}.

In species with female-biased reproductive sex ratios or extensive male-male competiton, high variance in male reproductive success can also reduce the $N_{e}$ of genes on the Y chromosome \citep{caballero1995,charlesworth2001,laporte2002,pool2007,ellegren2009}, suggesting that inferences about the effects of linked selection on the Y need to be distinguished from neutral models that include sex ratio variation as well as male-female differences in reproductive success. Because variance in male reproductive success reduces both Y-linked $N_{e}$ and autosomal $N_{e}$ \citep{kimura1964number,nomura2002effective}, evidence for this effect can be obtained by comparing levels of neutral variability on X and Y chromosomes, relative to values obtained for the autosomes. In particular, high variance in male reproductive success predicts a reduction in the Y/A diversity ratio, but an increase in the X/A ratio. Based on such comparisons, studies in humans have suggested examples of historical sex-differences in dispersal during human migration \citep{wilkins2006unraveling}, though it the inflated X/A ratio in humans has also been attributed to a historical excess of breeding females over males \citep{hammer2008sex} (but see \citep{bustamante2009,hammer2010,cotter2016genetic})

Despite widespread interest in determining the evolutionary factors affecting sex-linked variability \citep{ellegren2011,bachtrog2013NRG}, we still know very little about the influence of demography or selection on patterns of nucleotide diversity in more recently evolved sex chromosomes, such as in dioecious plants. We therefore don't yet know the time-scales over which processes such as linked selection on the Y chromosome are likely to be important. In humans, for example, estimates of Y-linked diversity are considerably lower than predicted under models of neutral evolution, and simulations have revealed that levels of Y-diversity are consistent with a combination of very low $N_{m}$ and strong purifying selection eliminating linked neutral variation \citep{Wilsonsayres2014}. However, given that human sex chromosomes evolved from autosomes \textasciitilde 200 million years ago (MYA), is an open question whether such strong effects of purifying selection should also apply to systems where sex chromosomes evolved \textit{do novo} from autosomes over much more recent evolutionary time (e.g., within the last 20 MYA as in the case of diecious plants \citep{charlesworth2015plant}). The effects of diversity loss due to purifying selection might on the one hand be expeted to be lower in such systems because of a shorter history of recombination suppression. Contrary to this intuition, however, simulations have shown that background selection and selective sweeps both have the strongest effects during the earliest stages of sex chromosome evolution, before Y-chromosomes have lost the majority of their genes \citep{bachtrog2008temporal}. Indeed, studies of Y-degeneration in the young plant sex chromosomes \textit{R. hastatulus} and \textit{Silene latifolia} have found signs of rapid Y chromosome degeneration \citep{hough2014,papadopulos2015,charlesworth2016plant}. In these cases, if the early stages of Y-chromosome evolution involved widespread Y-linked gene-inactivation (silencing), as predicted by \citep{orr1998adaptive}, and discovered in \textit{Drosophila mrianada} \citep{zhou2012chromosome}, then it is possible that the strength of linked selection would then be minimized such that the inferred gene loss was primarily driven by gene-silencing followed by neutral genetic drift \citep{bachtrog2013NRG}.

>>>>

One fundamental difference between the X and Y chromosomes is that there are 1/3 as many Y-linked gene copies as X-linked ones in a diploid population. Because of this, the effective population size ($N_{e}$) of the Y chromosome is expected to be 1/4 that of the autosomes, whereas the $N_{e}$ for the X chromosome should be 3/4 that of autosomes (assuming an equal number of reproducing females and males). The lowered $N_{e}$ of the Y chromosome implies that  deleterious mutations are more likely to rise to high frequency because of genetic drift \citep{nei1981genetic}. Moreover, this loss of recombination on the Y implies that genes on this chromosome will experience strong linkage disequilibrium, making them vulnerable to "Hill-Robertson Interference", in which the genetic associations between selected alleles and their genetic backgrounds cause a reduction in the efficacy of selection \citep{fisher1930genetical, muller1964relation, hill1966HReffect, mcvean2000}. This suggests that Y chromosomes should harbor a lower amount of neutral genetic variability than predicted based on the number of Y chromosomes in a population (proportional to the product of $N_{e}$ and the mutation rate, $\mu$, at equilibrium), though the extent to which such interference should affect chromosome-wide levels of nucleotide diversity is not known. Given that a large number of sites are probably under weak purifying selection (\X refs), the "interference" effect on Y chromosomes with suppressed recombination might alone cause a substantial diversity loss. Indeed, simulations suggest that the effects of HRI on 100,000 linked sites can reduce polymorphism levels by \textasciitilde 30 \% compared to neutral expectations \citep{mcvean2000}, suggesting that HRI is likely important for the evolution of Y chromosomes.

In species with female-biased reproductive sex ratios or extensive male-male competition, high variance in male reproductive success can also reduce the $N_{e}$ of genes on the Y chromosome \citep{caballero1995,charlesworth2001,laporte2002,pool2007,ellegren2009}, which suggests that inferences about the effects of HRI (including the effects of purifying and positive selection) on the Y need to be distinguished from neutral models that include sex ratio variation as well as sex differences in reproductive success. Because variance in male reproductive success reduces both Y-linked $N_{e}$ and autosomal $N_{e}$ \citep{kimura1964number,nomura2002effective}, evidence for this effect can be therefore obtained by comparing levels of neutral diversity on X and Y chromosomes relative to values on autosomes. For example, high variance in male reproductive success causes a reduction in the Y/A diversity ratio, but an increase in the X/A ratio. Based on such comparisons, studies in humans have suggested examples of historical sex-differences in dispersal during human migration \citep{wilkins2006unraveling}, though the inflated X/A ratio has also been attributed to a historical excess of breeding females over males \citep{hammer2008sex} (but see \citep{bustamante2009,hammer2010,cotter2016genetic})

Despite widespread interest in determining the evolutionary factors affecting neutral diversity on both X and Y chromosomes \citep{ellegren2011,bachtrog2013NRG}, we still know very little about the influence of demography or selection on patterns of sex-linked diversity in more recently evolved sex chromosomes, such as in dioecious plants. This information is important, however, as it highlights the relevant time scales over which effects such as linked selection are likely to be important. In humans, estimates of Y-linked diversity are considerably lower than predicted under models of neutral evolution, with simulations showing that current levels of Y-diversity are consistent with a combination of very low $N_{m}$ and a history of strong purifying selection \citep{Wilsonsayres2014}. However, given that human sex chromosomes evolved from autosomes \textasciitilde 200 million years ago (MYA), is not immediately obvious whether such strong effects of purifying selection on diversity should also apply to sex chromosomes that evolved \textit{do novo} from autosomes over much more recent evolutionary time (e.g., within the last 20 MYA in the case of most dioecious plants \citep{charlesworth2015plant}). The diversity loss due to purifying selection might on the one hand be expected to be lower in such systems due recombination having been suppressed on Y-chromosome for a shorter length of time. Contrary to this intuition, however, simulations have shown that two of the key types if HRI - background selection and selective sweeps - are both expected to have the strongest effects during the earliest stages of sex chromosome evolution, before Y-chromosomes have lost the majority of their genes \citep{bachtrog2008temporal}. Indeed, this is consistent with studies of Y-degeneration in young plant sex chromosomes that have found signs of rapid genetic degeneration \citep{hough2014,papadopulos2015,charlesworth2016plant}. In these cases, if the early stages of Y-chromosome evolution involved widespread Y-linked gene-inactivation (silencing), as predicted by \citep{orr1998adaptive}, and discovered on the neo-Y chromosome in \textit{Drosophila mrianada} \citep{zhou2012chromosome}, then it is possible that linked selection plays a less important role than expected in the early stages, with gene loss instead being primarily driven by gene-silencing followed by neutral genetic drift \citep{bachtrog2013NRG}.
